Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-07-08 paper/oopsla.tex
Submission

@@

\citet{DBLP:journals/pacmpl/JagadeesanJR20} note that the promising semantics
(\PS{}) \cite{DBLP:conf/popl/KangHLVD17} and related models
\citep{DBLP:conf/esop/JagadeesanPR10,DBLP:journals/pacmpl/ChakrabortyV19,Manson:2005:JMM:1047659.1040336},
fail to validate compositional reasoning of temporal properties.  Consider
their example \ref{oota4}:
\begin{align}
  \taglabel{oota4}
  \begin{gathered}
    \PW{y}{x}
    \PAR
    \PR{y}{r} \SEMI \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
    \PAR
    \PW{b}{1}
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx}{\DR{x}{1}}{}
        \event{wy}{\DW{y}{1}}{right=of rx}
        \po{rx}{wy}
        \event{ry}{\DR{y}{1}}{right=3em of wy} 
        \event{wx}{\DW{x}{1}}{right=of ry}
        \event{wz}{\DW{z}{1}}{right=of wx}
        \event{rb}{\DR{b}{1}}{right=of wz}
        \event{wb1}{\DW{b}{1}}{right=3em of rb}
        \po{ry}{wx}
        \rf{wb1}{rb}
        \rf{wy}{ry}
        \rf[out=-170,in=-10]{wx}{rx}
        \po{rb}{wz}
        \po[out=15,in=165]{ry}{wz}
      \end{tikzinline}}
  \end{gathered}  
\end{align}
Under \PS{}, this outcome is allowed by \emph{baiting} with the \texttt{else}
branch, then \emph{switching} to the \texttt{then} branch, based on a coin
flip $(\IF{b})$.\footnote{Call the threads \texttt{s}, \texttt{t}, and
  \texttt{u}.  To get the result in the promising semantics, first execute
  \texttt{u} to get message \texttt{<b:1@1>}.  Then \texttt{t} promises
  \texttt{<x:1@1>}, which it can fulfill by reading \texttt{b}$=$\texttt{0}.
  Then execute \texttt{s} to get message \texttt{<y:1@1>}.  Then execute
  \texttt{t}, reading \texttt{b}$=$\texttt{1} and \texttt{y}$=$\texttt{1} and
  fulfill the promise by writing \texttt{<x:1@1>}. The execution is exactly
  the same in our speculative semantics \cite{DBLP:conf/esop/JagadeesanPR10},
  removing timestamps and replacing the word \emph{promise} by
  \emph{speculation}.}

@@

\citet[Fig.~8]{DBLP:journals/toplas/Lochbihler13} notes that such violations
of temporal reasoning could result in violations of the Java security
architecture.

The difference between \PS{} and \PwT{} can be understood in terms of the
valid program transformations.  \PS{} allows reads to be introduced, with
subsequent case analysis (aka, if-closure) on the value read---effectively,
this can turn one read into two, with different conditional branches taken
for the two copies of the read.  For detailed examples, see
\cite{promising-ldrf}.

\PwT{} invalidates read introduction.  In return, \PwT{} enjoys
compositionality for temporal safety properties, as discussed in
\textsection\ref{sec:results} and
\cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20}.

This difference highlights the subtle tensions in relaxed memory models.
It is not possible to have everything one wants, thus, one is forced
to choose which optimizations and reasoning principles are most important.

@@

The substitutions collapse $\aLoc$ and $\aReg$, allowing local invariant
reasoning (\xLIR{}), as required by \jmm{} causality test case 1, discussed at the end of
\textsection\ref{sec:ex:control}.  Without Skolemizing it is necessary to
substitute $[\aLoc/\aReg]$, since the reverse substitution $[\aReg/\aLoc]$ is
useless when $\aReg$ is bound---compare with
\textsection\ref{sec:substitutions}.  As discussed below (\ref{p:downset}),
including this substitution affects the interaction of \xLIR{} and downset
closure.

@@

The proof of compilation to Arm in \jjr{} assumes that all internal reads can
be eliminated.
% Shortly after publication, \citet{anton}
% noticed a shortcoming of the implementation on \armeight{} in
% \jjr{\textsection 7}.  The proof given there assumes that all internal reads
% can be dropped.
However, this is not the case for acquiring reads.  For example, \jjr{}
disallows the following execution, where the final values of $x$ is $2$ and
the final value of $y$ is $2$.  This execution is allowed by \armeight{} and
\tso{}.
\begin{gather*}
  \PW{x}{2}\SEMI 
  \PR[\mACQ]{x}{r}\SEMI
  \PR{y}{s} \PAR
  \PW{y}{2}\SEMI
  \PW[\mREL]{x}{1}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{2}}{}
      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
      \event{c}{\DR{y}{0}}{right=of b}
      \event{d}{\DW{y}{2}}{right=2.5em of c}
      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
      \rf{a}{b}
      \sync{b}{c}
      \wk{c}{d}
      \sync{d}{e}
      \wk[out=-165,in=-15]{e}{a}
      % \rfi{a}{b}
      % \bob{b}{c}
      % \fre{c}{d}
      % \bob{d}{e}
      % \coe[out=-165,in=-15]{e}{a}
    \end{tikzinline}}
\end{gather*}
We discussed two approaches to this problem in \textsection\ref{sec:arm}.
% The solution we have adopted is to allow an acquiring read to be downgraded
% to a relaxed read when it is preceded (sequentially) by a relaxed write that
% could fulfill it.  This solution allows executions that are not allowed under
% \armeight{} since we do not insist that the local relaxed write is actually
% read from.  This may seem counterintuitive, but we don't see a local way to
% be more precise.

@@

Contrary to the claim, redundant read elimination fails for \jjr{}.
We discussed redundant read elimination in \textsection\ref{sec:semreg}.
Consider JMM Causality Test Case 2, which we discussed there.
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
  \PAR
  \PW{x}{y}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{1}}{right=of a1}
      \event{a3}{\DW{y}{1}}{right=of a2}
      \event{b1}{\DR{y}{1}}{right=3em of a3}
      \event{b2}{\DW{x}{1}}{right=of b1}
      \rf{a3}{b1}
      \po{b1}{b2}
      \rf[out=169,in=11]{b2}{a2}
      \rf[out=169,in=11]{b2}{a1}
    \end{tikzinline}}
\end{gather*}
Under the semantics of \jjr{}, we have
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{1}}{right=of a1}
      \event{a3}{1\EQ1\land1\EQ x \land x\EQ1 \land x=x\mid\DW{y}{1}}{right=of a2}
    \end{tikzinline}}
\end{gather*}
The precondition of $\DWP{y}{1}$ is \emph{not} a tautology, and therefore
redundant read elimination fails.
(It is a tautology in
\begin{math}
  \PR{x}{r}\SEMI
  \LET{s}{r}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
\end{math}.)
\jjr{\textsection3.1} incorrectly stated that the precondition of
$\DWP{y}{1}$ was $1\EQ1\land x\EQ x$.

>>>>

\section{Discussion}
\label{sec:discussion}

@@

Recently, \citet{promising-ldrf} showed that certain combinations of compiler
optimizations are inconsistent with local \drf{} guarantees.  All of the
examples that prove inconsistency have the same shape: they combine read
introduction and case analysis (aka, if-closure).  Effectively, this turns
one read into two, where different conditional branches can be taken for the
two copies of the read.  This is reminiscent of the type of \emph{bait and
  switch} behavior noted by \citet{DBLP:journals/pacmpl/JagadeesanJR20}: the
promising semantics (\PS{}) \cite{DBLP:conf/popl/KangHLVD17} and related
models
\citep{DBLP:conf/esop/JagadeesanPR10,DBLP:journals/pacmpl/ChakrabortyV19,Manson:2005:JMM:1047659.1040336},
fail to validate compositional reasoning of temporal properties.  Consider
their example \ref{oota4}:
\begin{align}
  \taglabel{oota4}
  \begin{gathered}
    \PW{y}{x}
    \PAR
    \PR{y}{r} \SEMI \PR{b}{s} \SEMI\IF{s}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
    \PAR
    \PW{b}{1}
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx}{\DR{x}{1}}{}
        \event{wy}{\DW{y}{1}}{right=of rx}
        \po{rx}{wy}
        \event{ry}{\DR{y}{1}}{right=3em of wy} 
        \event{wx}{\DW{x}{1}}{right=of ry}
        \event{wz}{\DW{z}{1}}{right=of wx}
        \event{rb}{\DR{b}{1}}{right=of wz}
        \event{wb1}{\DW{b}{1}}{right=3em of rb}
        \po{ry}{wx}
        \rf{wb1}{rb}
        \rf{wy}{ry}
        \rf[out=-170,in=-10]{wx}{rx}
        \po{rb}{wz}
        \po[out=15,in=165]{ry}{wz}
      \end{tikzinline}}
  \end{gathered}  
\end{align}
Under \PS{}, this outcome is allowed by \emph{baiting} with the \texttt{else}
branch, then \emph{switching} to the \texttt{then} branch, based on a coin
flip $(\IF{b})$.
% Using a combination of read introduction and case analysis,
% the middle thread can be rewritten to:
% \begin{displaymath}
%   \begin{array}{l}
%     \PR{y}{r} \SEMI \IF{b}\THEN\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \ELSE\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \FI
%   \end{array}
% \end{displaymath}
% \footnote{Call the threads \texttt{s}, \texttt{t}, and
%   \texttt{u}.  To get the result in the promising semantics, first execute
%   \texttt{u} to get message \texttt{<b:1@1>}.  Then \texttt{t} promises
%   \texttt{<x:1@1>}, which it can fulfill by reading \texttt{b}$=$\texttt{0}.
%   Then execute \texttt{s} to get message \texttt{<y:1@1>}.  Then execute
%   \texttt{t}, reading \texttt{b}$=$\texttt{1} and \texttt{y}$=$\texttt{1} and
%   fulfill the promise by writing \texttt{<x:1@1>}. The execution is exactly
%   the same in our speculative semantics \cite{DBLP:conf/esop/JagadeesanPR10},
%   removing timestamps and replacing the word \emph{promise} by
%   \emph{speculation}.}
% \citet[Fig.~8]{DBLP:journals/toplas/Lochbihler13} notes that such violations
% of temporal reasoning could result in violations of the Java security
% architecture.

@@

The examples in \cite{promising-ldrf} are more complex.  Here is one, dubbed
\textsc{ldrf-fail-ps}:
\begin{gather*}  
  %\taglabel{LDRF-Fail-PS}
  \begin{gathered}
  \IF{x}\THEN
    \PFADD{w}{}{1}\SEMI
    \PW{y}{1}\SEMI
    \PW{z}{1}
  \FI
  \PAR
  \IF{\BANG z}\THEN
    \PW{x}{1}
  \ELSE
    \IF{\BANG \PFADD{w}{}{1}}\THEN
      \PW{x}{\PR{y}{}}
    \FI
  \FI
    \\
    \hbox{\begin{tikzinlinesmall}[node distance=2em]
        \event{a1}{\DR{x}{1}}{}
        \event{a2}{\DR{w}{1}}{right=of a1}
        \event{a3}{\DW{w}{2}}{right=of a2}
        \event{a4}{\DW{y}{1}}{right=of a3}
        \event{a5}{\DW{z}{1}}{right=of a4}
        \event{b1}{\DR{z}{1}}{right=5em of a5}
        \event{b2}{\DR{w}{0}}{right=of b1}
        \event{b3}{\DW{w}{1}}{right=of b2}
        \event{b4}{\DR{y}{1}}{right=of b3}
        \event{b5}{\DW{x}{1}}{right=of b4}
        \rmw{a2}{a3}
        \po[out=15,in=165]{a1}{a3}
        \po[out=15,in=165]{a1}{a4}
        \po[out=15,in=165]{a1}{a5}        
        \rmw{b2}{b3}
        \po{b4}{b5}
        \po[out=15,in=165]{b2}{b5}        
        \po[out=15,in=165]{b1}{b3}
        \rf{a5}{b1}
        \rf[out=15,in=165]{a4}{b4}
        \rf[out=-165,in=-15]{b3}{a2}
        \rf[out=-165,in=-15]{b5}{a1}
      \end{tikzinlinesmall}}
  \end{gathered}
\end{gather*}
This execution is allowed by \PS{} and disallowed by \PwTmca{}.

The difference between \PS{} and \PwT{} can be understood in terms of the
valid program transformations.  \PS{} allows reads to be introduced, with
subsequent case analysis on the value read.  \PwT{} validates case analysis,
but invalidates read introduction.  In return, \PwT{} enjoys compositionality
for temporal safety properties, as discussed in \textsection\ref{sec:results}
and \cite[\textsection6]{DBLP:journals/pacmpl/JagadeesanJR20}.

\citet{promising-ldrf} show that by restricting \RMW{}-store reorderings, one
can establish local \drf{} for \PS{}.  We believe that no such restriction is
required for \PwT{}.

These differences highlight the subtle tensions between compiler optimizations
and program logics that are revealed by relaxed memory models.  It is not
possible to have everything one wants, thus, one is forced to choose which
optimizations and reasoning principles are most important.

@@

The substitutions collapse $\aLoc$ and $\aReg$, allowing local invariant
reasoning (\xLIR{}), as required by \jmm{} causality test case 1, discussed in
\textsection\ref{sec:q}.  Without Skolemizing it is necessary to
substitute $[\aLoc/\aReg]$, since the reverse substitution $[\aReg/\aLoc]$ is
useless when $\aReg$ is bound---compare with
\textsection\ref{sec:substitutions}.  As discussed below (\ref{p:downset}),
including this substitution affects the interaction of \xLIR{} and downset
closure.

@@

The proof of compilation to Arm in \jjr{} assumes that all internal reads can
be eliminated.
% Shortly after publication, \citet{anton}
% noticed a shortcoming of the implementation on \armeight{} in
% \jjr{\textsection 7}.  The proof given there assumes that all internal reads
% can be dropped.
However, this is not the case for acquiring reads.  For example, \jjr{}
disallows the following execution, where the final values of $x$ is $2$ and
the final value of $y$ is $2$.  This execution is allowed by \armeight{} and
\tso{}.
\begin{gather*}
  \PW{x}{2}\SEMI 
  \PR[\mACQ]{x}{r}\SEMI
  \PR{y}{s} \PAR
  \PW{y}{2}\SEMI
  \PW[\mREL]{x}{1}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a}{\DW{x}{2}}{}
      \raevent{b}{\DR[\mACQ]{x}{2}}{right=of a}
      \event{c}{\DR{y}{0}}{right=of b}
      \event{d}{\DW{y}{2}}{right=2.5em of c}
      \raevent{e}{\DW[\mREL]{x}{1}}{right=of d}
      \rf{a}{b}
      \sync{b}{c}
      \wk{c}{d}
      \sync{d}{e}
      \wk[out=-165,in=-15]{e}{a}
      % \rfi{a}{b}
      % \bob{b}{c}
      % \fre{c}{d}
      % \bob{d}{e}
      % \coe[out=-165,in=-15]{e}{a}
    \end{tikzinline}}
\end{gather*}
We discuss two approaches to this problem in \textsection\ref{sec:arm}.
% The solution we have adopted is to allow an acquiring read to be downgraded
% to a relaxed read when it is preceded (sequentially) by a relaxed write that
% could fulfill it.  This solution allows executions that are not allowed under
% \armeight{} since we do not insist that the local relaxed write is actually
% read from.  This may seem counterintuitive, but we don't see a local way to
% be more precise.

@@

Contrary to the claim, redundant read elimination fails for \jjr{}.
We discuss redundant read elimination in \textsection\ref{sec:semreg}.
Consider JMM Causality Test Case 2, which we describe there.
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
  \PAR
  \PW{x}{y}
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{1}}{right=of a1}
      \event{a3}{\DW{y}{1}}{right=of a2}
      \event{b1}{\DR{y}{1}}{right=3em of a3}
      \event{b2}{\DW{x}{1}}{right=of b1}
      \rf{a3}{b1}
      \po{b1}{b2}
      \rf[out=169,in=11]{b2}{a2}
      \rf[out=169,in=11]{b2}{a1}
    \end{tikzinline}}
\end{gather*}
Under the semantics of \jjr{}, we have
\begin{gather*}
  \PR{x}{r}\SEMI
  \PR{x}{s}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
  \\
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{a1}{\DR{x}{1}}{}
      \event{a2}{\DR{x}{1}}{right=of a1}
      \event{a3}{1\EQ1\land1\EQ x \land x\EQ1 \land x=x\mid\DW{y}{1}}{right=of a2}
    \end{tikzinline}}
\end{gather*}
The precondition of $\DWP{y}{1}$ is \emph{not} a tautology, and therefore
redundant read elimination fails.
(It is a tautology in
\begin{math}
  \PR{x}{r}\SEMI
  \LET{s}{r}\SEMI
  \IF{r{=}s}\THEN \PW{y}{1}\FI
\end{math}.)
\jjr{\textsection3.1} incorrectly stated that the precondition of
$\DWP{y}{1}$ was $1\EQ1\land x\EQ x$.

### 2021-07-08 paper/outro.tex
Submission

This paper is the first to present a direct denotational semantics for
sequential composition in a relaxed memory model which can be
efficiently compiled to modern CPUs. There is, as usual, more research
to be done.

@@

We've also not handled access elimination: store-forwarding and
dead-write-removal are unsound.  We expect that these can be validated by
allowing events with different labels to merge in some instances.

\PwTmca{1} is a simpler model than \PwTmca{2}, but requires fences on
acquiring reads for \armeight{}.  It would be illuminating to find out what
the performance penalty is for these fences.

In a similar vein, the promising semantics (\PS{}) validates read
introduction whereas \PwTmca{} does not.  As a result \PS{} admits behaviors
that break reasoning about temporal safety properties (see
\textsection\ref{sec:promising}).  Nonetheless, read introduction is
ubiquitous is some compilers.  It would be interesting to know if there is a
performance penalty for banning read introduction.

>>>>

This paper is the first to present a direct denotational semantics for
sequential composition in a relaxed memory model which can be
efficiently compiled to modern CPUs. We extract from this model a semantic
dependency relation and use it to build \PwTc, a solution to the
Out-of-Thin-Air problem in \cXI, and \PwTmca{}, a model for Java-like
languages.
% There is, as usual, more research to be done.

@@

% \PwTmca{} does not validate access elimination: store-forwarding and
% dead-write-removal are unsound.  We expect that these can be validated by
% allowing events with different labels to merge in some instances.

% \PwTmca{1} is a simpler model than \PwTmca{2}, but requires fences on
% acquiring reads for \armeight{}.  It would be illuminating to find out what
% the performance penalty is for these fences.

% In a similar vein,
\PwT{} does not validate read introduction, whereas speculative operational
semantics do.  Recent work shows a tension between read introduction and
compositional reasoning for temporal safety properties (see
\textsection\ref{sec:promising}).  Nonetheless, read introduction is
ubiquitous is some compilers.  It would be interesting to know if there is a
performance penalty for banning read introduction.
